Abstract

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Make minor changes to response letter style

September 16, 2024. <!-- TODO: Update date -->

@@

Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (se description of argument `SE.type`) for a detailed description of how this package computes the standard errors.

@@

Our work is completely theoretical and thus uses a purely deductive method.
Therefore, we do not have any hypothesis to test, but the results of the formal derivations that we present, which address the objective described at the end of our introduction.

@@

Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
(That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
To correct this, we have modified the immediately preceding paragraph to the following:

@@

The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
We have added a clarification in this respect to the end of the paragraph, which now reads:

@@

In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
We have now added a paragraph to our introduction to try to explain this fact intuitively.
Please refer to our answer to the editor on p. 3 of this letter for more details on this addition.

Regarding the person parameters, the interpretation is more complicated:
If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e., its covariance is an identity matrix).
This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.

Our first version tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p. 14, lines 15-22, or p. <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).

@@

This is the expression of a vector represented in a basis, in this case, representing that the vector can be represented in each of the two bases interchangeably.
Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:

@@

This is a mistake on our side, we are afraid to admit, product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
We have changed the term by the mathematically appropriate "interchangeably" in our new version.

@@

We think that our word choice has been unfortunate here.
What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
This is implicit in the definition of the test space itself (see p. 7 line 8 in our original submission, or p. <!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen, which does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction.

@@

As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p. <!-- TODO: Complete --> in this letter).
In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", and not when we represent the attributes.

@@

However, in our manuscript, we intend to go one step further and show how the latent space covariance affects how the item parameters are interpreted.
We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor on the effect each latent trait has on the response probability function.
As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.
Therefore, the discrimination parameter values should be considered as being "parallel" to the latent space axis.
In other words, each discrimination parameter value is itself the projection of the item vector length onto its corresponding axis.

>>>>

June 2, 2025. <!-- TODO: Update date -->

@@

Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (please see description of argument `SE.type`) for a detailed description of how this package computes the standard errors.

@@

Our work is entirely theoretical and thus employs a purely deductive method.
Consequently, we do not have any hypothesis to test; rather, we present the results of our formal derivations, which address the objective described at the end of our introduction.

@@

<!-- # TODO: Equation references change? -->
Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
(That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
To correct this, we have modified the immediately preceding paragraph to the following:

@@

The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
We have added a clarification to this at the end of the paragraph, which now reads:

@@

In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
We have now added a paragraph to our introduction to try to explain this fact intuitively.
Please refer to our answer to the editor on pp. 3-4 of this letter for more details on this addition.

Regarding the person parameters, the interpretation is more complicated:
If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e., its covariance is an identity matrix).
This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p. 14, lines 15-22, or p. <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).

@@

This is the expression of a vector represented in a basis, in this case, stating that the vector can be represented in each of the two bases interchangeably.
Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:

@@

This is a mistake on our side, we are afraid to admit, the product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
We have changed the term to the mathematically appropriate "interchangeably" in our new version.

@@

We think that our word choice has been unfortunate here.
What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
This is implicit in the definition of the test space itself (see p. 7 line 8 in our original submission, or p. <!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
This basis does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.

@@

As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p. <!-- TODO: Complete --> in this letter).
In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", regardless of how we represent the attributes.

@@

However, in our manuscript, we intend to go one step further and show how the latent space covariance affects how the item parameters are interpreted.
We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor of the effect each latent trait has on the response probability function.
As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.
Therefore, the discrimination parameter values should be considered as being "parallel" to the latent space axis.
In other words, each discrimination parameter value is itself the projection of the item vector length onto its corresponding axis.
